<!--
%\VignetteEngine{knitr::docco_linear}
%\VignetteIndexEntry{Outcrossing populations}
-->

# Outcrossing  populations

```{r, globalsetup, echo=FALSE, results='hide', cache=FALSE}
#opts_chunk$set(cache=TRUE, autodep=TRUE)
```

```{r, echo=FALSE, results='hide'}
library(onemap)
```

The following example is intended to show the usage of `OneMap`
functions for linkage mapping in outcrossing (non-inbred) populations.
With basic knowledge of R syntax, one should have no problems using
it. If you are not familiar with R, we recommend reading the vignette
[Introduction to R](http://htmlpreview.github.com/?https://github.com/augusto-garcia/onemap/blob/master/vignettes_html/Introduction_R.html).

Hopefully these examples will be clear enough to help any user to
understand its functionality and start using it. You do not need to be
an expert in R to build your linkage map, but some concepts are
necessary and will help you through the process.

## Creating the data file

This step may be quite difficult, because the data file is not very
simple and some errors can occur while reading it. The input file
format is similar to that used by MAPMAKER/EXP (Lander et al., 1987),
so experienced users of genetic analysis software should be already
familiar with this scenario.

Basically, the input file is a text file, where the first line indicates the cross type and the second line provides information about the number of individuals, the number of markers, the presence of physical marker locations and the presence of phenotypic data. The third line contains sample IDs. Then, the genotype information is included separately for each marker. The
character `*` indicates the beginning of information input for a new
marker, followed by the marker name. Next, there is a code indicating the marker type, according to Wu's et al. (2002a)
notation. Actually, it is recommended to check Wu's et al. (2002a)
paper before using `OneMap`.

Marker types must be one of the following: `A.1`, `A.2`, `A.3`, `A.4`,
`B1.5`, `B2.6`, `B3.7`, `C.8`, `D1.9`, `D1.10`, `D1.11`, `D1.12`,
`D1.13`, `D2.14`, `D2.15`, `D2.16`, `D2.17` or `D2.18`, each one
corresponding to a row of the following table:

|     |       |           | Parent         |                |    Offspring   |             |
|-----|-------|:---------:|:--------------:|:--------------:|:--------------:|:-----------:|
|     |       | Crosstype | Cross          | Observed bands | Observed bands | Segregation |
| $A$ |       |         1 | $ab \times cd$ | $ab \times cd$ | $ac, ad, bc, bd$ | $1:1:1:1$ |
|     |       |         2 | $ab \times ac$ | $ab \times ac$ | $a, ac, ba, bc$ | $1:1:1:1$|
|     |       |         3 | $ab \times co$ | $ab \times c$  | $ac, a, bc,b$  | $1:1:1:1$ |
|     |       |         4 | $ao \times bo$ | $a  \times  b$ | $ab, a, b, o$  | $1:1:1:1$ |
|     |       |           |                |                |       |  |
| $B$ | $B_1$ |         5 | $ab \times ao$ | $ab \times a$  | $ab, 2a,b$ | $1:2:1$ |
|     | $B_2$ |         6 | $ao \times ab$ | $a \times ab$  | $ab,2a,b$  | $1:2:1$ |
|     | $B_3$ |         7 | $ab \times ab$ | $ab \times ab$ | $a, 2ab,b$ | $1:2:1$ |
|     |       |           |                |                |       |  |
| $C$ |       |         8 | $ao \times ao$ | $a \times a$   | $3a, o$        |  $3:1$      |
|     |       |           |                |                |       |  |
| $D$ | $D_1$ |         9 | $ab \times cc$ | $ab \times c$  | $ac, bc$       |  $1:1$      |
|     |       |        10 | $ab \times aa$ | $ab \times a$  | $a, ab$        |  $1:1$      |
|     |       |        11 | $ab \times oo$ | $ab \times o$  | $a, b$         |  $1:1$      |
|     |       |        12 | $bo \times aa$ | $b \times a$   | $ab, a$        |  $1:1$      |
|     |       |        13 | $ao \times oo$ | $a \times o$   | $a, o$         |  $1:1$      |
|     | $D_2$ |        14 | $cc \times ab$ | $c \times ab$  | $ac, bc$       |  $1:1$      |
|     |       |        15 | $aa \times ab$ | $a \times ab$  | $a, ab$        |  $1:1$      |
|     |       |        16 | $oo \times ab$ | $o \times ab$  | $a, b$         |  $1:1$      |
|     |       |        17 | $aa \times bo$ | $a \times b$   | $ab, a$        |  $1:1$      |
|     |       |        18 | $oo \times ao$ | $o \times a$   | $a, o$         |  $1:1$      |

Letters `A`, `B`, `C` and `D` indicate the segregation type (_i.e._,
`1:1:1:1`, `1:2:1`, `3:1` or `1:1`, respectively), while the number
after the dot (_e.g._, `A.1`) indicates the observed bands in the
offspring. The paper cited above gives details with respect to marker
types; we will not discuss them here, but it is easy to see that each
marker is classified based on the band patterns of parents and
progeny.

Finally, after each marker name, comes the genotype data for the
segregating population. The coding for marker genotypes used by
`OneMap` is also the same one proposed by Wu et al. (2002a) and the
possible values vary according to the specific marker type. Missing
data are indicated with the character `-` (minus sign) and an empty space separates the information for each individual. Phenotype information, if present, follows genotypic data with a similar structure. Details are found in the help of function `read_onemap`.

Here is an example of such file for 10 individuals and 5 markers (the three zeros in the second line indicate that there is no chromosome information, physical position information or phenotypic data, respectively). It
is very similar to a MAPMAKER/EXP file, but has additional information
about the crosstype.

```
data type outcross
10 5 0 0 0
I1 I2 I3 I4 I5 I6 I7 I8 I9 I10
*M1 B3.7 ab ab - ab b ab ab - ab b
*M2 D2.18 o - a a - o a - o o
*M3 D1.13 o a a o o - a o a o
*M4 A.4 ab b - ab a b ab b - a
*M5 D2.18 a a o - o o a o o o
```

Notice that once the marker type is identified, no variations of
symbols presented on the table for the **observed bands** is allowed.
For example, for `A.1`, only `ac`, `ad`, `bc` and `bd` genotypes are
expected (plus missing values). **We notice in FAQs that this is a
common mistake made by users, so please be careful**.

The input file must be saved in text format, with extensions like
`.raw`. It is a good idea to open the text file called
`example_out.raw` (available with `OneMap` and saved in the directory
you installed it) to see how this file should be. You can see where
`OneMap` is installed using the command

```{r, eval=FALSE}
system.file(package = "onemap")
``` 

## Importing data

Once the input file is created, data can be loaded and saved into an R
object. The function used to import data is named `read_onemap`. Its usage is quite simple:
 
```{r, eval=FALSE}
example_out <- read_onemap("C:/workingdirectory", "example_out.raw")
```

```{r, eval=FALSE, echo=FALSE}
example_out <- read_onemap(system.file("extdata", package = "onemap"), "example_out.raw")
```

The first argument is the directory where the input file is located,
so modify it accordingly. The second one is the data file name. In
this example, an object named `example_out` was created. If you leave
the argument `dir` blank, the file will be loaded from your `working
directory`.

You can change the working directory in R using function `setwd()` or
in the toolbar clicking `File -> Change dir`. If you set your working
directory to the one containing the input file, you can just type:

```{r, eval=FALSE}
example_out <- read_onemap(file = "example_out.raw")
``` 

If no error has occurred, a message will display some basic
information about the data, such as number of individuals and number
of markers:

```{r, echo=FALSE}
example_out <- read_onemap(system.file("extdata", package = "onemap"), "example_out.raw")
``` 

Because this particular data set is distributed along with the
package, as an alternative you can load it typing

```{r}
data(example_out)
``` 

Loading the data creates an object of class `onemap`, which will
further be used in the analysis. R command `print` recognizes
objects of this class. Thus, if you type

```{r, eval=FALSE}
example_out
```

you will see some information about the object:

```{r, echo=FALSE}
example_out
```

## Estimating two-point recombination fractions

To start the analysis, the first step is estimating the recombination
fraction between all pairs of markers, using two-point tests:

```{r, twopoint, results='hide'}
twopts <- rf_2pts(example_out)
``` 

The function `rf_2pts` uses as default values of LOD Score `3` and
maximum recombination fraction `0.50`.

Different values for the criteria can be chosen using:

```{r, linkcriteria, results='hide'}
twopts <- rf_2pts(example_out, LOD = 3, max.rf = 0.4)
``` 

Although two-point tests were implemented in C language, which is usually much
faster than R, this step can take quite some time, depending on the
number of markers involved and their segregation type, because all
combinations will be estimated and tested. Besides, the results use a
lot of memory and a rather powerful computer is needed.

When the two-point analysis is finished, an object of class `rf_2pts`
created. Typing

```{r, eval=FALSE}
twopts
```

will show a message with the criteria used in the analysis and some
other information:

```{r, echo=FALSE}
twopts
``` 

If you want to see the results for given markers, say `M1`
and `M3`, the command is:

```{r}
print(twopts, c("M1", "M3"))
``` 

Each line corresponds to a possible linkage phase. `CC` denotes coupling
phase in both parents, `CR` and `RC` denote coupling phase in parent 1
and 2, respectively, and repulsion in the other, and `RR`
denotes repulsion phase in both parents. Value `rf` is the maximum
likelihood estimate of the recombination fraction, with its corresponding
LOD Score.

## Assigning markers to linkage groups

Once the recombination fractions and linkage phases for all pairs of
markers have been estimated and tested, markers can be assigned to
linkage groups. To do this, first use the function `make_seq` to
create a sequence with the markers you want to assign:

```{r}
mark_all <- make_seq(twopts, "all")
```

The function `make_seq` is used to create sequences from objects of
several kinds, as will be seen along this tutorial. Here, the object
is of class `rf_2pts` and the second argument specifies which markers
one wants to use. In this example, the argument `"all"` indicates that
all markers will be analyzed. If one wants to use only a subset of
markers, say `M1` and `M2`, the option will be `c(1,2)`. These numbers
refer to the lines where markers are located on the data file. Because
the identification of the markers can be cumbersome, one should use
the function `marker type` to see their numbers, names and types:


```{r} 
marker_type(mark_all)
``` 

The grouping step is very simple and can be done by using the function
`group`:

```{r}
LGs <- group(mark_all)
```

For this function, optional arguments are `LOD` and `max.rf`, which
define thresholds to be used when assigning markers to linkage groups.
If none provided (default), criteria previously defined for the object
`twopts` are used.

The previous command generates an object of class `group` and the
command `print` for such object has two options. If you type:

```{r, eval=FALSE}
LGs
``` 

you will get detailed information about the groups, that is, all linkage
groups will be printed, displaying the names of markers in each one of
them.

```{r, echo=FALSE}
LGs
``` 

However, in case you just want to see some basic information (such as
the number of groups, number of linked markers, etc), use

```{r}
print(LGs, detailed = FALSE)
``` 

You can notice that all markers are linked to some linkage group. If
the LOD Score threshold is changed to a higher value, some markers are
kept unassigned:

```{r}
LGs <- group(mark_all, LOD = 6)
LGs
``` 

Changing back to the previous criteria, now setting the maximum recombination fraction to 0.40:
```{r}
LGs <- group(mark_all, LOD = 3, max.rf = 0.4)
LGs
``` 

## Genetic mapping of linkage group 3


Once marker assignment to linkage groups is finished, the mapping step
can take place. First of all, you must set the mapping function that
should be used to display the genetic map throughout the analysis. You
can choose between `Kosambi` or `Haldane` mapping functions. To use
Haldane, type

```{r, eval=FALSE}
set_map_fun(type = "haldane")
```

To use Kosambi

```{r, eval=FALSE}
set_map_fun(type = "kosambi")
```

Now, you must define which linkage group will be mapped. In another
words, a linkage group must be `extracted` from the object of class
`group`, in order to be mapped. For simplicity, we will start here
with the smallest one, which is linkage group 3. This can be easily
done using the following code:

```{r}
LG3 <- make_seq(LGs, 3)
```

The first argument (`LGs`) is an object of class `group` and the
second is a number indicating which linkage group will be extracted,
according to the results stored in object `LGs`. The object `LG3`,
generated by function `make_seq`, is of class `sequence`, showing that
this function can be used with several types of objects.

If you type

```{r, eval=FALSE}
LG3
``` 

you will see which markers are comprised in the sequence, and also
that no parameters have been estimated so far.

```{r, echo=FALSE}
LG3
``` 

To order these markers, one can use a two-point based algorithm such
as Seriation (Buetow and Chakravarti, 1987), Rapid Chain Delineation
(Doerge, 1996), Recombination Counting and Ordering (Van Os et al.,
2005) and Unidirectional Growth (Tan and Fu, 2006):
  
```{r,  results='hide'}
LG3_ser <- seriation(LG3)
LG3_rcd <- rcd(LG3)
LG3_rec <- record(LG3)
LG3_ug <- ug(LG3)
``` 

In this case, all algorithms provided the same results (results not shown).

To order by comparing all possible orders (exhaustive search), the
function `compare` can be used:

```{r,  results='hide'}
LG3_comp <- compare(LG3)
``` 

This algorithm can take some time to run, depending on marker types in
the linkage group. In the example, `LG3` contains one marker of type
D1 and one of type D2, besides one marker segregating in 3:1 fashion
(type C). Thus, although the number of possible orders is relatively
small (60), for each order there are various possible combinations of
linkage phases. Also, the convergence of the EM algorithm takes
considerably more time, because markers of type C are not very
informative.

The first argument to the `compare` function is an object of class
`sequence` (the extracted group `LG3`), and the object generated by
this function is of class `compare`.

To see the results of the previous step, type

```{r}
LG3_comp
``` 

Remember that for outcrossing populations, one needs to estimate
marker order and also linkage phases between markers for a given
order. However, because two point analysis provides information
about linkage phases, this information is taken into consideration in
the `compare` function, reducing the number of combinations to be
evaluated. If a given linkage phase has LOD greater than 0.005 in the two point analysis, we assume that this phase is very unlikely and so does not need to be evaluated in the multipoint procedure used by `compare`. We did extensive simulations which showed that this is a good procedure.

By default, `OneMap` stores 50 orders, which may or may not be unique.
The value of `LOD` refers to the overall LOD Score, considering all
orders tested. `Nested LOD` refers to LOD Scores within a given order, that is, scores for different combinations of linkage phases for the same
marker order.

For example, order 1 has the largest value of log-likelihood and,
therefore, its LOD Score is zero for a given combination of linkage
phases (CC, CC, RR, RR). For this same order and other linkage phases, LOD Score is -5.20. Analyzing the results for order 2, notice that its highest LOD Score is very close to zero, indicating that this order is also quite plausible. Notice also that `Nested LOD` will always contain at least one zero value, corresponding to the best combination of phases for markers in a given order. Due to the information provided by two-point analysis, not all combinations are tested and that is the reason why the number of Nested LOD values is different for each order.

Unless one has some biological information, it is a good idea to
choose the order with the highest likelihood. The final map can then
be obtained with the command

```{r}
LG3_final <- make_seq(LG3_comp, 1, 1)
``` 

The first argument is the object of class `compare`. The second argument indicates which order is chosen: 1 is for the order with highest likelihood, 2 is for the second best, and so on. The third argument indicates which combination of phases is chosen for a given order: 1 also means the combination with highest likelihood among all combinations of phases (based on Nested LOD).

For simplicity, these values are defaults, so typing

```{r}
LG3_final <- make_seq(LG3_comp)
``` 

will have the same effect.

To see the final map, type

```{r}
LG3_final
``` 

At the leftmost position, marker names are displayed. `Position` shows
the cumulative distance using the Kosambi mapping function. Finally,
`Parent 1` and `Parent 2` show the diplotypes of both parents, that
is, the combination in which alleles are arranged in the chromosomes, given the estimated linkage phase. Notation is the same as that used by Wu et al. (2002a). Details about how ordering algorithms can be chosen and used are presented by Mollinari et al. (2009).

## Genetic mapping of linkage group 2

Now let us map the markers in linkage group number 2.
 
 Again, `extract` that group from the object `LGs`:
 
```{r}
LG2 <- make_seq(LGs, 2)
LG2
``` 

Note that there are 10 markers in this group, so it is infeasible to
use the `compare` function with all of them because it will take a very
long time to proceed.

First, use `rcd` to get a preliminary order estimate:

```{r}
LG2_rcd <- rcd(LG2)
LG2_rcd
```

Use the `marker_type` function to check the segregation types of all
markers in this group:

```{r, , echo=TRUE}
marker_type(LG2)
``` 

Based on their segregation types and distribution on the preliminary
map, markers `M4`, `M23`, `M19`, `M20` and `M24` are the most
informative ones (type `A` is better, followed by type `B`). So,
let us create a framework of ordered markers using `compare` for the
most informative ones:

```{r, results='hide'}
LG2_init <- make_seq(twopts, c(4, 23, 19, 20, 24))
LG2_comp <- compare(LG2_init)
``` 

```{r}
LG2_comp
```

Now, the first argument to `make_seq` is an object of class `rf_2pts`,
and the second argument is a vector of integers, specifying which
molecular markers comprise the sequence.

Select the best order:

```{r}
LG2_frame <- make_seq(LG2_comp)
``` 

Next, let us try to map the remaining markers, one at a time. Because
there are more markers of type D1 than D2, the latter will be tried
later. Starting with `M9`:

```{r, results='hide'}
LG2_extend <- try_seq(LG2_frame, 9)
```

```{r}
LG2_extend
``` 

Based on the LOD Scores, marker M9 is probably better located between
markers M23 and M24. However, the `*` symbol indicates that more than
one linkage phase is possible. Detailed results can be seen with

```{r}
print(LG2_extend, 5)
```

The second argument indicates the position where to place the marker.
Note that the first allele arrangement is the most likely one.

Also, we can obtain some useful diagnostic graphics using the argument
`draw.try=TRUE` when using function `try_seq`:

```{r, echo=TRUE}
LG2_extend <- try_seq(LG2_frame, 9, draw.try = TRUE)
``` 

The top figure represents the new genetic map obtained with the
insertion of marker 9 between markers `M23` and `M24` (most likely
position). The left bottom figure represents the frame map `M24 - M23 -
M4 - M19 - M20` on the x-axis and the LOD Scores of the linkage maps
obtained with the insertion of marker 9 at the beginning, between
markers and at the end of the frame map. The red triangle indicates
the most likely position, where marker 9 is supposed to be
placed. The right bottom figure is the recombination fraction matrix
based on a color scale using the function `rf_graph_table`. The
diagnostic graphics show an almost monotonic recombination fraction
matrix (the values are bigger as their distance from diagonal
increases). This pattern is typical of well ordered linkage groups. We can
see that the position between markers `M23` and `M24` is the most
likely one for positioning marker `M9`.

Finally, the best order can be obtained with:

```{r}
LG2_frame <- make_seq(LG2_extend, 5, 1)
```

When using `make_seq` with an object of class `try`, the second
argument is the position on the map (according to the scale on the
right of the output) and the last argument indicates linkage phases
(defaults to 1, higher nested LOD).

It should be pointed out that the framework created by the function
`compare` (with `M20`, `M4`, `M19`, `M23` and `M24`) could be in
reverse order (`M24`, `M23`, `M19`, `M4` and `M20`) and still represent the
same map. Thus, positioning of markers with the `try_seq` command can be
different in your computer. For example, here marker `M9` was better
placed at position 5; however if you obtain a reversed order, marker
`M9` would be better placed in position 2. In both cases the best
position is between markers `M24` and `M23`.

Adding other markers, one by one (output not shown):

```{r, results='hide'}
LG2_extend <- try_seq(LG2_frame, 29)
LG2_frame <- make_seq(LG2_extend, 7)
LG2_extend <- try_seq(LG2_frame, 27)
LG2_frame <- make_seq(LG2_extend, 1)
LG2_extend <- try_seq(LG2_frame, 16)
LG2_frame <- make_seq(LG2_extend, 2)
LG2_extend <- try_seq(LG2_frame, 21)
LG2_final <- make_seq(LG2_extend, 6)
``` 

The process of adding markers sequentially can be automated with the
use of function `order_seq`.
  
```{r}
LG2_ord <- order_seq(LG2, n.init = 5, THRES = 3, draw.try = FALSE)
``` 

Basically, this function automates what the `try_seq` function does,
using some pre-defined rules. In the function, `n.init = 5` means that
five markers (the most informative ones) will be used in the `compare`
step; `THRES = 3` indicates that the `try_seq` step will only add
markers to the sequence which can be mapped with LOD Score greater
than 3; `draw.try = TRUE` will display a diagnostic graphic for each
`try_seq` step (in this example, we set this to `draw.try = FALSE`),
which should be combined with `wait = 1`, that indicates the minimum
time interval in seconds to display the diagnostic graphic.

*NOTE*: Although very useful, this function can be misleading,
 specially if there are not many fully informative markers, so use it
 carefully. Results can vary between multiple runs on the same markers, of course.
 
Check the final order:

```{r}
LG2_ord
```

Note that markers `16` and `29` could not be safely mapped to a single
position (`LOD Score > THRES` in absolute value). The output displays
the `safe` order and the most likely positions for markers not mapped,
where `***` indicates the most likely position and `*` corresponds to
other plausible positions.

To get the safe order (_i.e._, without markers 16 and 29), use

```{r}
LG2_safe <- make_seq(LG2_ord, "safe")
```

and to get the order with all markers, use

```{r}
LG2_all <- make_seq(LG2_ord, "force")
LG2_all
```

Notice that, for this linkage group, the `forced` map obtained with
`order_seq` is the same as that obtained with `compare` plus
`try_seq`, but *this is not always the case*.

The `order_seq` function can also perform two rounds of the `try_seq`
algorithms, first using `THRES` and then `THRES - 1` as threshold.
This generally results in safe orders with more markers mapped, but
may take longer to run. To do this use the `touchdown` option:

```{r, results='hide'}
LG2_ord <- order_seq(LG2, n.init = 5, THRES = 3, touchdown = TRUE)
```

```{r}
LG2_ord
```

For this particular sequence, the `touchdown` step could map marker M29, but this depends on the specific dataset.

Finally, to check for alternative orders (because we did not use
exhaustive search), use the `ripple_seq` function:

```{r}
ripple_seq(LG2_all, ws = 4, LOD = 3)
``` 

We should do this to any of the orders we found, either using
`try_seq` or `order_seq`. Here, we choose `LG2_all` for didactic purposes only. The second argument, `ws = 4`, means that
subsets (windows) of four markers will be permutated sequentially
(`4!` orders for each window), to search for other plausible orders.
The `LOD` argument means that only orders with LOD Score smaller than
3 will be printed.

The output shows sequences of four numbers, because `ws = 4`. They are
followed by an `OK` if there is no alternative order with LOD
Score smaller than `LOD = 3` in absolute value, or by a list of
alternative orders. In the example, the first and last sequences showed
alternative orders with LOD smaller than `LOD = 3`. However, the best order was the original one (`LOD = 0.00`).

If there were an alternative order more likely than the original, one
should check the difference between these orders (and linkage phases)
and change it using, for example, functions `drop_marker` and
`try_seq`, or by typing the new order. You can use `$seq.num` and
`$seq.phases` after the name of the sequence (for example,
`LG2_all$seq.num` and `LG2_all$seq.phases`) to obtain the original
order and linkage phases, make the necessary changes (by copying and
pasting) and then use the function `map` to reestimate the genetic map for the new order.

Here, the `ripple_seq` function showed that the final order obtained
is indeed the best for this linkage group. The map can then be printed
using
    
```{r}
LG2_all
``` 

## Genetic mapping of linkage group 1

Finally, linkage group 1 (the largest one) will be analyzed. Extract markers:

```{r}
LG1 <- make_seq(LGs, 1)
``` 

Construct the linkage map, by automatically using try algorithm:

```{r, results='hide'}
LG1_ord <- order_seq(LG1, n.init = 6, touchdown = TRUE)
```

```{r}
LG1_ord
```

Notice that the second round of `try_seq` added markers `M5` and `M25`.

Now, get the order with all markers:

```{r}
(LG1_final <- make_seq(LG1_ord, "force"))
``` 

Check the final map:
```{r}
ripple_seq(LG1_final)
``` 

No better order was observed.

Print it:
```{r}
LG1_final
```

As an option, different algorithms to order markers could be applied: 

```{r}
LG1_ser <- seriation(LG1)
LG1_rcd <- rcd(LG1)
LG1_rec <- record(LG1)
LG1_ug  <- ug(LG1)
``` 

There are some differences between the results. Seriation did not
provide good results in this case. See Mollinari et al. (2009) for an
evaluation of these methods.
  
## Map estimation for an arbitrary order

If, for any reason, one wants to estimate parameters for a given
linkage map (_e.g._, for other orders on published papers), it is
possible to define a sequence and use the `map` function. For example,
for markers `M30`, `M12`, `M3`, `M14` and `M2`, in this order, use
  
```{r}
any_seq <- make_seq(twopts, c(30, 12, 3, 14, 2))
(any_seq_map <- map(any_seq))
``` 

This is a subset of the first linkage group. When used this way, the `map`
function searches for the best combination of phases between markers
and prints the results.

Furthermore, a sequence can also have user-defined linkage phases. The
next example shows (incorrect) phases used for the same order of
markers:

```{r}
any_seq <- make_seq(twopts, c(30, 12, 3, 14, 2), phase = c(4, 1, 4, 3))
(any_seq_map <- map(any_seq))
``` 

If one needs to add or drop markers from a predefined sequence,
functions `add_marker` and `drop_marker` can be used. For example, to
add markers 4 to 8 to `any.seq`.

```{r}
(any_seq <- add_marker(any_seq, 4:8))
```

Removing markers 3, 4, 5, 12 and 30 from `any.seq`:

```{r}
(any_seq <- drop_marker(any_seq, c(3, 4, 5, 12, 30)))
```

After that, the map needs to be re-estimated.

## Plotting the recombination fraction matrix

For a given sequence, it is possible to plot the recombination
fraction matrix and LOD Scores based on a color scale using the
function `rf_graph_table`. This matrix can provide some useful diagnostics about the map.
  
For example, using the function `group` with `LOD = 2.5`:

```{r}
(LGs <- group(mark_all, LOD = 2.5))
``` 

Due to the small value used for the LOD Score (2.5, not adequate and
resulting in false positives), markers from different groups were
placed together.

Ordering markers (results not shown):

```{r, results='hide'}
LG_err <- make_seq(LGs, 2)
LG_err_ord <- order_seq(LG_err)
```

The map using option `force`:

```{r, eval=FALSE}
(LG_err_map <- make_seq(LG_err_ord, "force"))
``` 

```{r, echo=FALSE}
LG_err_map <- map(make_seq(twopts, c(27, 16, 20, 4, 19, 21, 23, 9, 24, 29, 22, 7, 18, 8, 13), phase = c(1, 1, 2, 4, 1, 3, 1, 2, 3, 3, 2, 1, 4, 4)))
LG_err_map
``` 

A careful examination of the results shows that there are problems on the map. This can be done by plotting the recombination fraction matrix:

```{r}
rf_graph_table(LG_err_map)
```

```{r,  echo=FALSE}
rf_graph_table(LG_err_map, inter = FALSE)
``` 

The recombination fractions are plotted below the diagonal and the LOD Scores are plotted above the diagonal. The color scale varies from red (small distances or big LODs) to dark blue. This color scale follows the `rainbow` color palette with `start` argument equal to 0 and `end` argument equal to 0.65. White cells indicate combinations of markers for which the recombination fractions cannot be estimated (D1 and D1).

Clicking on the cell corresponding to two markers (off the secondary diagonal), you can see some information about them. For example, clicking on the cell corresponding to markers `M4` and `M19` you can see their names, types (`A.4` and `B1.5`), recombination fraction (`rf = 0.02281`) and LOD Scores for each possible linkage phase. Clicking on a cell on the diagonal, some information about the corresponding marker is shown, including percent of missing data. This is quite useful in helping to interpret the results.

Looking at the matrix, it is possible to see two groups: one with
markers from `LG2` (`M27`, `M16`, `M20`, `M4`, `M19`, `M21`, `M23`,
`M9`, `M24` and `M29`) and other with markers from `LG3` (`M22`, `M7`,
`M18`, `M8`, and `M13`). There is a gap between markers `M22` and
`M29` (`rf = 0.4594`). At this position, the group should be divided,
that is, a higher LOD Score should be used. Notice that these two
groups were placed together due to a false linkage (false positive)
detected between markers `M4` and `M22` (LOD Score `2.9`), because of not using an appropriate LOD threshold (a more conservative
value).

The `rf_graph_table` can also be used to check the order of markers
based on the monotonicity of the matrix: as we get away from the
secondary diagonal, the recombination fraction values should increase.

## Drawing the genetic map

Once all linkage groups were obtained, we can draw a simple map using
the function `draw_map`. We can draw a genetic map for all linkage
groups:

```{r, echo=TRUE, fig=TRUE}
maps <- list(LG1_final, LG2_final, LG3_final)
draw_map(maps, names = TRUE, grid = TRUE, cex.mrk = 0.7)
``` 

For a specific linkage group:

```{r, width = 9.5, height = 9.5}
draw_map(LG1_final, names = TRUE, grid = TRUE, cex.mrk = 0.7)
``` 

It is obvious that function `draw_map` draws a very simple graphic
representation of the genetic map. But once the distances and the
linkage phases are estimated, better map figures can be drawn by the
user with any appropriate software. There are several free software
that can be used, such as `MapChart` (Voorrips, 2002).


## References

Buetow, K. H., Chakravarti, A. Multipoint gene mapping using
seriation. I. General methods. **_American Journal of Human
Genetics_** 41, 180-188, 1987.

Doerge, R.W. Constructing genetic maps by rapid chain delineation.
**_Journal of Agricultural Genomics_** 2, 1996.

Mollinari, M., Margarido, G. R. A., Vencovsky, R. and Garcia, A. A. F.
Evaluation of algorithms used to order markers on genetics maps.
**_Heredity_** 103, 494-502, 2009.

Tan, Y., Fu, Y. A novel method for estimating linkage maps.
**_Genetics_** 173, 2383-2390, 2006.

Van Os H, Stam P, Visser R.G.F., Van Eck H.J. RECORD: a novel method
for ordering loci on a genetic linkage map. **_Theor Appl Genet_**
112, 30-40, 2005.

Voorrips, R.E. MapChart: software for the graphical presentation of
linkage maps and QTLs. **_Journal of Heredity_** 93, 77-78, 2002.

Wu, R., Ma, C.X., Painter, I. and Zeng, Z.-B. Simultaneous maximum
likelihood estimation of linkage and linkage phases in outcrossing
species. **_Theoretical Population Biology_** 61, 349-363, 2002a.

Wu, R., Ma, C.-X., Wu, S. S. and Zeng, Z.-B. Linkage mapping of
sex-specific differences. **_Genetical Research_** 79, 85-96, 2002b.
